Avoiding false discoveries: Revisiting an Alzheimer’s disease snRNA-Seq dataset

Version 1

Abstract

Abstract

Arising From

Mathys, H. et al. Nature (2019). https://doi.org/10.1038/s41586-019-1195-2 Mathys et al., conducted the first single-nucleus RNA-Seq study (snRNA-Seq) of Alzheimer’s disease (AD)1. The authors profiled the transcriptomes of approximately 80,000 cells from the prefrontal cortex, collected from 48 individuals – 24 of which presented with varying degrees of AD pathology. With bulk RNA-Seq, changes in gene expression across cell types can be lost, potentially masking the differentially expressed genes (DEGs) across different cell types. Through the use of single-cell techniques, the authors benefitted from increased resolution with the potential to uncover cell type-specific DEGs in AD for the first time2. However, there were limitations in both their data processing and quality control and their differential expression analysis. Here, we correct these issues with best-practice approaches to snRNA-Seq processing and differential expression, resulting 892 times fewer differentially expressed genes at a false discovery rate (FDR) of 0.05.

Main

Our questions of Mathys et al.,1 focus around their data processing and their differential expression analysis. Firstly, in relation to their processing approach, the authors discussed the high percentages of mitochondrial reads and low number of reads per cell present in their data. This is indicative of low cell quality3 however, we believe the authors’ quality control (QC) approach did not capture all of these low quality cells. Moreover, the authors did not integrate the cells from different individuals to account for batch effects. As the field has matured since the author’s work was published, dataset integration has become a standard step in single-cell RNA-Seq protocols4. To address these issues, we used scFlow4 to reprocess the author’s data. This pipeline included the removal of empty droplets, nuclei with low read counts and doublets followed by embedding and integration of cells from separate samples and cell typing. See Supplementary Note 1 for a detailed explanation of these steps. Re-processing resulted in 50,831 cells passing QC, approximately 20,000 less the authors’ post-processing set. Additionally, we found different cell type proportions to the authors (Fig. 1a) which could be due to accounting for batch effects.

a shows the proportion of cells left after quality control (QC) from the author’s original work (Mathys et al.) and our reanalysis (Our analysis). b, c highlights the log2 fold change and -log10 false discovery rate (FDR) of the differentially expressed genes. In c, we have marked an FDR of 5×10−7, dashed grey line, to highlight how small the p-values from Mathys et al.’s analysis are. d, e, f, g show the Pearson correlation between the cell counts after QC and the number of DEGs identified. The different cell types are astrocytes (Astro), excitatory neurons (Exc), inhibitory neurons (Inh), microglia (Micro), oligodendrocytes (Oligo) and oligodendrocyte precursor cells (OPC).

Our second question of Mathys et al.,1 is their differential expression approach. The authors conducted a differential expression analysis between the controls and the patients with AD pathology, concentrating on six neuronal and glial cell types; excitatory neurons, inhibitory neurons, astrocytes, microglia, oligodendrocytes and oligodendrocyte precursor cells, derived from the Allen Brain Atlas5. They performed downstream analysis on their identified DEGs and investigated some of the most compelling genes in more detail. Therefore, all findings put forward by their paper were based upon the validity of their differential expression approach. However, for this approach, the authors conducted a two-part, cell and patient level analysis. The cell-level analysis took each cell as an independent replicate and the results of which were compared for consistency in directionality and rank of their DEGs against their patient level analysis, a Poisson mixed model. The authors identified 1,031 DEGs using this combinatorial approach – DEGs requiring an FDR <0.01 in the cell-level and an FDR<0.05 in the patient level analysis. It is important to note that this cell-level differential expression approach, also known as pseudoreplication, over-estimates the confidence in DEGs due to the statistical dependence between cells from the same patient not being considered6,7,8. When we inspect all DEGs identified at an FDR of 0.05 from the authors’ cell-level analysis, this number increases to 14,274. Pseudobulk differential expression (DE) analysis has recently been proven to give optimal performance compared to both mixed models and pseudoreplication approaches6,9. It aggregates counts to individuals thus accounting for the dependence between an individual’s cells.

Here, we apply a pseudobulk DE approach, sum aggregation and edgeR LRT10, to the reprocessed data. We found 16 unique DEGs when considering the six cell types used by the authors (Supplementary Table 1). This was 892 times fewer DEGs than that reported originally at an FDR of 0.05. When we compare these DEGs, we can see that the absolute log2 fold change (LFC) of our DEGs is over 20 times larger than the authors’; median LFC of 3.33 and 0.16, despite the authors’ DEGs having an FDR score 76,000 times smaller; median FDR of 2.89×10−7 and 0.02 (Fig. 1b-c). We can show that this stark contrast in these DEGs is just an artefact of the authors taking cells as independent replicates by considering the Pearson correlation between the number of DEGs found and the cell counts (Fig. 1d-f). There is a near perfect correlation for the authors DEGs, whether we consider the DEGs from the pseudoreplication analysis (Fig. 1d) or the 1,031 genes from the authors’ combinatorial approach (Fig. 1e) which is not present in our re-analysis (Fig. 1f).

Although it has been proven that pseudoreplication approaches result in false positives by artificially inflating the confidence from non-independent samples, we wanted to investigate the effect of the approach on the authors’ dataset. We ran the same cell-level analysis approach – a Wilcoxon rank-sum test and FDR multiple-testing correction, 100 times whilst randomly permuting the patient identifiers (Fig. 1g). We would expect not to find any differential expressed genes with this approach given the random mixing of case and control patients. However, this pseudoreplication approach consistently found high numbers of DEGs and we observe the same correlation between the number of cells and number of DEGs as with the authors results. As a result, we conclude that integrating this pseudoreplication approach with a mixed model like the authors proposed just artificially inflates the test confidence for a random sample of the genes resulting in more false discoveries in cell types with bigger counts.

Conclusion

In conclusion, the authors’ analysis has been highly influential in the field with numerous studies undertaken based on their results, something we show has uncertain foundations. However, we would like to highlight that the use of pseudoreplication in neuroscience research is not isolated to the author’s work, others have used this approach11,12,13 and their results should be similarly scrutinised. Here, we provide our processed count matrix with metadata and also, the DEGs identified using an independently validated, differential expression approach so that other researchers can use this rich dataset without relying on pseudoreplication approaches. While the number of DEGs found here are significantly lower, much greater confidence can be had that these are AD relevant genes. The low number of DEGs found may also cause concern given the sample size and cost of collection and sequencing of such datasets. However, the increasing number of snRNA-Seq studies being conducted for AD, creates the opportunity to conduct differential meta-analyses to increase power. Further work is required in the field to develop methods to conduct such analysis, integrating studies and accounting for their the heterogeneity, similar to that which has been done for bulk RNA-Seq14. Some such approaches have already been made in COVID-19 research which could be leveraged for neurodegenerative diseases15.

Supporting information

eLife Assessment

This work is useful as it highlights the importance of data analysis strategies in influencing outcomes during differential gene expression testing. While the manuscript has the potential to enhance awareness regarding data analysis choices in the community, its value could be further enhanced by providing a more comprehensive comparison of alternative methods and discussing the potential differences in preprocessing, such as scFLOW. The current analysis, although insightful, appears incomplete in addressing these aspects.

Reviewer #1 (Public Review):

Murphy, Fancy and Skene performed a reanalysis of snRNA-seq data from Alzheimer Disease (AD) patients and healthy controls published previously by Mathys et al. (2019), arriving at the conclusion that many of the transcriptional differences described in the original publication were false positives. This was achieved by revising the strategy for both quality control and differential expression analysis. I believe the authors' intention was to show the results of their reanalysis not as a criticism of the original paper (which can hardly be faulted for their strategy which was state-of-the-art at the time and indeed they took extra measures attempting to ensure the reliability of their results), but primarily to raise awareness and provide recommendations for rigorous analysis of sc/snRNA-seq data for future studies.

STRENGTHS:

The authors demonstrate that the choice of data analysis strategy can have a vast impact on the results of a study, which in itself may not be obvious to many researchers.

The authors apply a pseudobulk-based differential expression analysis strategy (essentially, adding up counts from all cells per individual and comparing those counts with standard RNA-seq differential expression tests), which is (a) in line with latest community recommendations, (b) different from the "default options" in most popular scRNA-seq analysis suites, and (c) explains the vastly different number of DEGs identified by the authors and the original publication. The recommendation of this approach together with a detailed assessment of the DEGs found by both methodologies could be a useful finding for the research community. Unfortunately, it is currently not fully substantiated and is confounded with concurrent changes in QC measures (see weaknesses).

The authors show a correlation between the number of DEGs and the number of cells assessed, which indicates a methodological shortcoming of the original paper's approach (actually, the authors of the original paper already acknowledged that the lesser number of DEGs for rare cell types was a technical artefact). To be educational for the reader it would be important to provide more information about the DEGs that were "found" and those that were "lost". Given vast inter-individual heterogeneity in humans, it is likely that the study was underpowered to find weaker differences using the pseudobulks (Fig. 1B shows that only genes with more than 4-fold change were found "significant").

All code and data used in this study are publicly available to the readers.

WEAKNESSES:

The authors interpret the fact that they found fewer DEGs with their method than the original paper as a good thing by making the assumption that all genes that were not found were false positives. However, they do not prove this, and it is likely that at least some genes were not found due to a lack of statistical power and not because they were actually "incorrect". The original paper also performed independent validations of some genes that were not found here.

I am concerned that the only DEGs found by the authors are in the rare cell types, foremost the rare microglia (see Fig. 1f). It is unclear to me how many cells the pseudo-bulk counts were based on for these cells types, but it seems that (a) there were few and (b) there were quite few reads per cells. If both are the case, the pseudobulk counts for these cell populations might be rather noisy and the DEG results are liable to outliers with extreme fold changes.

The authors claim they improved the quality control of the dataset. While I do not think they did anything wrong per se, the authors offer no objective metric to assess this putative improvement. This is another major weakness of the paper as it confounds the results of the improved (?) differential analysis strategy and dilutes the results. I detail this weakness in the two following points:

Removing low-quality cells: The authors apply a new QC procedure resulting in the removal of some 20k more cells than in the original publication. They state "we believe the authors' quality control (QC) approach did not capture all of these low quality cells" (l. 26). While all the QC metrics used are very sensible, it is unclear whether they are indeed "better". For instance, removal with a mitochondrial count of <5% seems harsh and might account for a large proportion of additional cells filtered out in comparison to the original analysis. There is no blanket "correct cutoff" for this percentage. For instance, the "classic" Seurat tutorial https://satijalab.org/seurat/articles/pbmc3k_tutorial.html uses the 5% threshold chosen by the authors, an MAD-based selection of cutoff arrived at 8% here https://www.sc-best-practices.org/preprocessing_visualization/quality_control.html, another "best practices" guide choses by default 10% https://bioconductor.org/books/3.17/OSCA.basic/quality-control.html#quality-control-discarded, etc. Generally, the % of mitochondrial reads varies a lot between datasets. As far as I can tell, the original paper did not use a fixed threshold but instead used a clustering approach to identify cells with an "abnormally high" mitochondrial read fraction. That also seems reasonable. Overall, I cannot assess whether the new QC is really more appropriate than the original analysis and the authors do not provide any evidence in favor of their strategy.

Batch correction: "Dataset integration has become a standard step in single-cell RNA-Seq protocols" (l. 29). While it is true that many authors now choose to perform an integration step as part of their analysis workflow, this is by no means uncontroversial as there is a risk of "over-integration" and loss of true biological differences. Also, there are many different methods for dataset integration out there, which will all have different results. More importantly, the authors go on "we found different cell type proportions to the authors (Fig. 1a) which could be due to accounting for batch effects" but offer no support for the claim that the batch effects are indeed related to the observed differences. An alternative explanation would be a selective loss/gain of certain cell types during quality control. The original paper stated concerns about losing certain cell types (microglia, which do not seem to be differentially abundant in the original paper / new analysis).

Relevant literature is incompletely cited. Instead of referring to reviews of best practices and benchmarks comparing methods for batch correction and or differential analysis, the authors only refer to their own previous work.

Due to a lack of comparison with other methods and due to the fact that the author's methodology was only applied to a single dataset, the paper presents merely a case study, which could be useful but falls short of providing a general recommendation for a best practice workflow.

APPRAISAL:

The manuscript could help to increase awareness of data analysis choices in the community, but only if the superiority of the methodology was clearly demonstrated. The recommended pseudobulk differential expression approach along with the indication of drastic differences that this might have on the results is the main output of the current manuscript, but it is difficult to assess unequivocally how this influenced the results because the differential analysis comes after QC and cell type annotation, which have also been changed in comparison to the original publication. In my opinion, the purpose of the paper might be better served by focusing on the DE strategy without changing QC and instead detailing where/how DEGs were gained/lost and supporting whether these were false positives.

Reviewer #2 (Public Review):

Summary: This paper takes on the important topic of preprocessing of single cell/nuclei RNA-seq prior to testing for differential gene expression. However, the manuscript has a number of critical weaknesses.

Strengths: This is an important topic and a key dataset for illustration.

Weaknesses: A major contribution is the use of the authors' own inhouse pipeline for data preparation (scFLOW), but this software is unpublished since 2021 and consequently not yet refereed. It isn't reasonable to take this pipeline as being validated in the field.

The authors claim that Mathys' analysis didn't use batch correction prior to analysis and claim that such processing is routine in the field, but the only citation they give is to the above-mentioned scFLOW. Batch correction for DEG analysis isn't the field standard, for example, Bryois et al. (2022) PMID: 35915177 doesn't perform batch correction. Whether or not to do such preprocessing is certainly arguable, but the authors need to argue it, not presuppose it.

The authors spend considerable effort in discounting the pseudoreplication analysis of Mathys. It is well understood that this analysis yields a lot of false positives, but Mathys only used this approach for removing genes, not as a valid test in and of itself. They also worry that the significant findings in Mathys' paper are influenced by the number of cells of each type. I'm sure it is since power is a function of sample size, but is this a bad thing? It seems odd that their approach is not influenced by sample size.

Author Response

The following is the authors’ response to the original reviews.

Response to Reviewers

To whom it may concern, Thank you for your constructive feedback on our manuscript. I appreciate the time and effort that you and the reviewers have dedicated to providing your valuable feedback. We are grateful to the reviewers for their insightful comments and suggestions for our paper. I have been able to incorporate changes to reflect the majority of these suggestions provided. I have updated the analysis scripts (at https://github.com/neurogenomics/reanalysis_Mathys_2019) and have listed these changes in blue below:

eLife assessment:

This work is useful as it highlights the importance of data analysis strategies in influencing outcomes during differential gene expression testing. While the manuscript has the potential to enhance awareness regarding data analysis choices in the community, its value could be further enhanced by providing a more comprehensive comparison of alternative methods and discussing the potential differences in preprocessing, such as scFLOW. The current analysis, although insightful, appears incomplete in addressing these aspects.

We thank the reviewing editors for this note. We agree that the differences in preprocessing will affect the results and conceal which step in our reanalysis resulted in the discrepancies we noted. To address this, we have split out our reanalysis into two separate parts - In the main body of the text we discuss the differences resulting from just changing the differential expression approach where we use the same processed data as the authors to enable a fair comparison. Secondly, we still provide the reprocessed data and perform differential expression analysis on it and discuss the cause and impact the differences in the processing steps made to the results.

Reviewer 1:

I think readers would be interested to learn more about the genes that were found "significant" by the original paper but sorted out by the authors. Did they just fall short of the cutoffs? If so, how many more samples would have been required to ascertain significance? This would yield a recommendation for future studies and an overall more positive/productive spirit to the manuscript. On the other hand, I suspect a fraction of DEGs were false positives due to differences in the proportions of cells from different individuals compared to the original analysis. Which percentage of DEGs does this apply to? Again, this would raise awareness of the issue and support the use of pseudobulk approaches.

To investigate the relationship between the genes and how they differ across our analysis we have added a correlation analysis between our different DE approaches (using the same processed data), see paragraph 5 in the manuscript and supplementary table 3. In short, we find that there is a high correlation in the genes’ fold change values across our pseudobulk analysis and the author’s pseudoreplication analysis on the same dataset (pearson R of 0.87 for an adjusted p-value of 0.05) which is somewhat expected given the DE approaches are applied to the same dataset. However, the p-values, which pertain to the likelihood that a gene’s expressional changes is related to the case/control differences in AD, and resulting DEGs vary considerably due to the artificially inflated confidence of the author’s approach (Fig. 1c-e). Despite there being a correlation between the pseudoreplciation and pseudobulk approaches here, we do not think it makes sense to consider how many more samples would have been required to ascertain significance. The differences in results between the two approaches is not negatable with sample size as many DEGs identified by pseudoreplication will be false positives as highlighted in previous work1,2,3,4. However, perhaps we are misinterpreting the reviewer, who may have meant a power analysis which we have not conducted. Such an undertaking would require analysing a multitude of snRNA-Seq of large sample sizes to garner a confident estimate for power calculations based on pseudobulk approaches. Although we agree with the reviewer that this would be beneficial to the field, we do not believe it is in scope for this work. On the reviewer’s note regarding a fraction of DEGs being false positives due to differences in the proportions of cells from different individuals compared to the original analysis - We have analysed the same processed data the authors used to negate the differences caused by the differing processing steps. We thank the reviewer for this suggestion. We also give more insight into the cause of these differences, namely on filtering our nuclei with large proportions of mitochondrial reads and discuss their effect in paragraph 3 (also see Supplementary Figure 2).

Given there are only a few DEGs, it would be good to show more data about these genes to allow better assessment of the robustness of the results, i.e., boxplots of the pseudobulk counts in the compared groups and perhaps heatmaps of the raw counts prior to aggregation. This could rule out concerns about outliers affecting the results.

In Supplementary Figure 3, we have added boxplots of the sum pseudobulked, trimmed mean of M-values (TMM) normalised counts for three of our identified DEGs (b) and three of the authors’ DEGs which they discuss in their manuscript (a) to show the differences in counts across AD pathology and controls for these genes. We hope this gives some insight into the transcriptional changes highlighted by the differing approaches. In our opinion, there is a clear difference in the transcriptional signal in the genes identified from pseudobulk which is not present for the genes identified from the authors approach.

Overall, I believe the paper would deliver a clearer message by mainlining the QC from the original study and only changing the DE analysis. However, if keeping the part about QC/batch correction:

Assess to which degree changes in cell type proportion are indeed due to batch correction (as suggested in the text) and not filtering by looking at the annotated cell types in the original publication and those in your analysis.

Also perform the analysis without changing QC and state the # of DEGs in both cases, to at least allow some disentanglement of the effect of different steps of the analysis.

Please state the number of cells removed by each QC step in the supplementary note.

We thank the reviewer for this suggestion. We agree with performing the DE analysis on the same processed data as the original authors and have split out our reanalysis into two separate parts, primarily focussing on the discrepancies caused by the choice of differential expression (DE) approach. By splitting our analysis in this manner, we can identify the substantial differences in results caused by differing the DE approach in the study. Secondly, we can see how differences in preprocessing affects the DE results in isolation too – see paragraph 8 but in short, the fold change correlation between pseudobulk DE analyses on the reprocessed data vs authors processed data only had a moderate correlation (Pearson R of 0.57).

In regards to the number of cells removed by each QC step, we have added an aggregated view for all samples in supplementary table 3 and also give the full statistics per sample in our Github repository: https://github.com/neurogenomics/reanalysis_Mathys_2019. Moreover, we investigated the root cause in the differences in nuclei numbers, uncovering filtering down to mitochondrial read proportions as the main culprit (Supplementary Figure 2).

I recommend the authors read the following papers, assess whether their methodology agrees with them, and add citations as appropriate to support statements made in the manuscript.

We thank the reviewer for this comprehensive list. We have updated our manuscript and supplementary file and main text throughout to cite many of these where appropriate. We believe this helps add context to our decisions for the differing tools and approaches used as part of the processing pipeline with scFlow and the differential expression approach.

I believe the authors' intention was to show the results of their reanalysis not as a criticism of the original paper (which can hardly be faulted for their strategy which was state-of-the-art at the time and indeed they took extra measures attempting to ensure the reliability of their results), but primarily to raise awareness and provide recommendations for rigorous analysis of sc/snRNA-seq data for future studies.

We thank the reviewer for this note, this was exactly our intent. Furthermore, we are based in a dementia research institute and our aim is to ensure that ensure that the Alzheimer’s disease research field does not focus on spuriously identified genes.We have updated the text of the manuscript (start paragraph 2) to explicitly state this so our message is not misconstrued.

In my opinion, the purpose of the paper might be better served by focusing on the DE strategy without changing QC and instead detailing where/how DEGs were gained/lost and supporting whether these were false positives.

We agree that the differences in preprocessing will affect the results and conceal which step in our reanalysis resulted in the discrepancies we noted. To address this, we have split out our reanalysis into two separate parts - In the main body of the text we discuss the differences resulting from just changing the differential expression approach where we use the same processed data as the authors to enable a fair comparison. Secondly, we still provide the reprocessed data and perform differential expression analysis on it and discuss the impact the differences in the processing steps made to the results. As previously mentioned, we have also added further investigation into the DEGs identified, looking at the correlation across the differing approaches and plotting the counts for selected genes.

For instance, removal with a mitochondrial count of <5% seems harsh and might account for a large proportion of additional cells filtered out in comparison to the original analysis. There is no blanket "correct cutoff" for this percentage. For instance, the "classic" Seurat tutorial https://satijalab.org/seurat/articles/pbmc3k_tutorial.html uses the 5% threshold chosen by the authors, an MAD-based selection of cutoff arrived at 8% here https://www.sc-best-practices.org/preprocessing_visualization/quality_control.html, another "best practices" guide choses by default 10% https://bioconductor.org/books/3.17/OSCA.basic/quality-control.html#quality-control-discarded, etc. Generally, the % of mitochondrial reads varies a lot between datasets.

Apologies, the 5% cut-off was a misprint – the actual cut-off used was 10% which, as the reviewer notes, is on the higher side of what is recommended. We have updated our manuscript to rectify this mistake and discuss the differences in the number of cells caused by the two approaches to mitochondrial filtering in the manuscript (paragraph 3). We found that over 16,000 nuclei that were removed in our QC pipeline were kept by the author’s (Supplementary Fig. 2), explaining the discrepancy in the number of nuclei after QC. Based on Supplementary Fig. 2, it is clear the author’s approach was ineffective at removing nuclei with high proportions of mitochondrial reads which is indicative of cell death5,6. We hope this alleviates the reviewer’s concerns around our alternative processing approach. Moreover, as mentioned, we swapped to compare the differences by DE approaches on the same data to avoid any effect by this.

Reviewer 2:

The paper would be better if the authors merged this work with the scFLOW paper so that they can justify their analysis pipeline and show it in an influential dataset.

We thank the reviewer for this note. We would like to clarify that the purpose of our work was not to show the scFlow analysis pipeline on an influential dataset but rather to raise awareness and provide recommendations for rigorous analysis of single-cell and single-nucleus RNA-Seq data (sc/snRNA-Seq) for future studies and to help redirect the focus of the Alzheimer’s disease research field away from possible spuriously identified genes. We have updated our manuscript text to highlight this (see start paragraph 2). Furthermore, we are aware our original approach reprocessing the data with scFlow will affect the results and conceal which step in our reanalysis resulted in the discrepancies we noted. Thus, we have split out our reanalysis into two separate parts - In the main body of the text we discuss the differences resulting from just changing the differential expression approach where we use the same processed data as the authors to enable a fair comparison. Secondly, we still provide the reprocessed data so that the community can benefit from it and perform differential expression analysis on it and discuss the impact the differences in the processing steps made to the results. We have also added further references supporting the choice of steps and tools used in scFlow in the supplementary text which should address the reviewer’s concerns about justifying the analysis pipeline. Moreover, we identified the cause of the nuclei count differences caused by the two processing approaches, namely on filtering our nuclei with large proportions of mitochondrial reads and discuss their effect in paragraph 3 (also see Supplementary Figure 2).

A major contribution is the use of the authors' own inhouse pipeline for data preparation (scFLOW), but this software is unpublished since 2021 and consequently not yet refereed. It isn't reasonable to take this pipeline as being validated in the field.

We believe our answer to the previous point addresses these concerns - We have added references supporting the choice of steps and tools used in scFlow in the supplementary text which should address the reviewer’s concerns about justifying the analysis pipeline. Moreover, as a result of the pipeline we identified that 16,000 of the nuclei kept by the authors are likely of low quality and indicative of cell death with high mitochondrial read proportions5,6.

They also worry that the significant findings in Mathys' paper are influenced by the number of cells of each type. I'm sure it is since power is a function of sample size, but is this a bad thing? It seems odd that their approach is not influenced by sample size.

We thank the reviewer for highlighting this point. As they noted, we conclude that the original authors number of DEGs is just a product of the number of cells. However, the reviewer states that ‘It seems odd that their approach is not influenced by sample size’. An increase in the number of cells is not an increase in sample size since these cells are not independent from one another - they come from the same sample. Therefore, an increase in the number of cells should not result in an increase in the number of DEGs whereas an increase in the number of samples would. This point is the major issue with pseudoreplication approaches which over-estimate the confidence when performing differential expression due to the statistical dependence between cells from the same patient not being considered. See these references for more information on this point1,2,7,8. We have added a discussion of this point to our manuscript in paragraph 6.

Moreover, recent work has established that the genetic risk for Alzheimer’s disease acts primarily via microglia9,10. Thus, it would be reasonable to expect that the majority of large effect size DEGs identified would be found in this cell type. This is what we found with our pseudobulk differential expression approach – 96% of all DEGs were in microglia. We have updated the text of our manuscript (paragraph 5) to highlight this last point.

References

1. Murphy, A. E. & Skene, N. G. A balanced measure shows superior performance of pseudobulk methods in single-cell RNA-sequencing analysis. Nat. Commun. 13, 7851 (2022).

2. Squair, J. W. et al. Confronting false discoveries in single-cell differential expression. Nat. Commun. 12, 5692 (2021).

3. Crowell, H. L. et al. muscat detects subpopulation-specific state transitions from multi-sample multi-condition single-cell transcriptomics data. Nat. Commun. 11, 6077 (2020).

4. Soneson, C. & Robinson, M. D. Bias, robustness and scalability in single-cell differential expression analysis. Nat. Methods 15, 255–261 (2018).

5. Ilicic, T. et al. Classification of low quality cells from single-cell RNA-seq data. Genome Biol. 17, 29 (2016).

6. Heumos, L. et al. Best practices for single-cell analysis across modalities. Nat. Rev. Genet. 24, 550–572 (2023).

7. Zimmerman, K. D., Espeland, M. A. & Langefeld, C. D. A practical solution to pseudoreplication bias in single-cell studies. Nat. Commun. 12, 738 (2021).

8. Lazic, S. E. The problem of pseudoreplication in neuroscientific studies: is it affecting your analysis? BMC Neurosci. 11, 5 (2010).

9. Skene, N. G. & Grant, S. G. N. Identification of Vulnerable Cell Types in Major Brain Disorders Using Single Cell Transcriptomes and Expression Weighted Cell Type Enrichment. Front. Neurosci. 0, (2016).

10. McQuade, A. & Blurton-Jones, M. Microglia in Alzheimer’s disease: Exploring how genetics and phenotype influence risk. J. Mol. Biol. 431, 1805–1817 (2019).

Version 2

Abstract

Abstract

Mathys et al., conducted the first single-nucleus RNA-Seq study (snRNA-Seq) of Alzheimer’s disease (AD)1. The authors profiled the transcriptomes of approximately 80,000 cells from the prefrontal cortex, collected from 48 individuals – 24 of which presented with varying degrees of AD pathology. With bulk RNA-Seq, changes in gene expression across cell types can be lost, potentially masking the differentially expressed genes (DEGs) across different cell types. Through the use of single-cell techniques, the authors benefitted from increased resolution with the potential to uncover cell type-specific DEGs in AD for the first time2. However, there were limitations in both their data processing and quality control and their differential expression analysis. Here, we correct these issues and use best-practice approaches to snRNA-Seq differential expression, resulting 549 times fewer differentially expressed genes at a false discovery rate (FDR) of 0.05.

Main

Mathys et al.’s1 data processing and quality control strategy for their single-nucleus RNA-Seq (snRNA-Seq) data was state-of-the-art at this time. Furthermore, the authors took extra measures in an attempt to ensure the reliability of their results. Our re-analysis is not a criticism of this but endeavours to raise awareness and provide recommendations for rigorous analysis of single-cell and single-nucleus RNA-Seq data (sc/snRNA-Seq) for future studies. Most importantly, we aim to ensure that the Alzheimer’s disease research field does not focus on spuriously identified genes. To this end, our questions of Mathys et al.,1 focus around their data processing and their differential expression analysis. Firstly, in relation to their processing approach, the authors discussed the high percentages of mitochondrial reads and low number of reads per cell present in their data. This is indicative of low cell quality3 however, we believe the authors’ quality control (QC) approach may not capture all of these low quality cells. Moreover, the authors did not integrate the cells from different individuals to account for batch effects. As the field has matured since the author’s work was published, dataset integration has become a common step in single-cell RNA-Seq protocols and is recommended to remove confounding sources of variation4,5,6. To gain advantage of these recent approaches, we used scFlow7 to reprocess the author’s data. This pipeline included the removal of empty droplets, nuclei with low read counts and doublets followed by embedding and integration of cells from separate samples and cell typing. ScFlow combines best practice approaches for processing sc/snRNA-Seq datasets, see Supplementary Note 1 for a detailed explanation of these steps. Re-processing resulted in 50,831 cells passing QC, approximately 20,000 less the authors’ post-processing set with differing cell type proportions (Supplementary Fig. 1).

In regards to data quality, it is worth noting that over 99% of nuclei had less than 200 genes expressed (Supplementary Table 3). While this QC step was not unique to our reprocessing, the authors made the same exclusion in their analysis1, it highlights the relatively low quality of the data which may be attributable to the early stage of snRNA-Seq technology of the time. For example, Brase et al.’s recent study of snRNA-Seq of autosomal dominant AD8 used a more stringent cut-off for the minimum number of expressed genes and still kept 27% (122 times more) of the assayed cells after all QC steps. Moreover, as mentioned, the authors discussed the high percentages of mitochondrial reads in their data. The differences in approaches to filtering based on the proportion mitochondrial reads accounts for the notable discrepancy in the number of nuclei after QC between our approach and the authors. Our approach used a 10% cut-off for the proportion of mitochondrial reads in a nuclei, as set out in Amezquita et al.’s best-practice guidelines5, which is less stringent than Seurat’s guidelines (5%)9 or that from Heumos et al. (8% from a median absolute deviations based cut-off selection)4. Conversely, the authors filtered out high mitochondrial read nuclei based on clusters from their t-SNE projection of the data1. Even at our lenient cut-off, over 16,000 nuclei that were removed in our QC pipeline were kept by the author’s (Supplementary Fig. 2), explaining the discrepancy in the number of nuclei after QC. Based on Supplementary Fig. 2, it is clear the author’s approach was ineffective at removing nuclei with high proportions of mitochondrial reads which is indicative of cell death3,4. We have made the data from our alternative processing approach publicly available (through Synapse: https://doi.org/10.7303/syn51758062.1) so researchers can utilise this resource free of low quality nuclei.

Our second question of Mathys et al.,1 is their differential expression approach. The authors conducted a differential expression analysis between the controls and the patients with AD pathology, concentrating on six neuronal and glial cell types; excitatory neurons, inhibitory neurons, astrocytes, microglia, oligodendrocytes and oligodendrocyte precursor cells, derived from the Allen Brain Atlas10. They performed downstream analysis on their identified DEGs and investigated some of the most compelling genes in more detail. Therefore, all findings put forward by their paper were based upon the validity of their differential expression approach. However, for this approach, the authors conducted a two-part, cell and patient level analysis. The cell-level analysis took each cell as an independent replicate and the results of which were compared for consistency in directionality and rank of their DEGs against their patient level analysis, a Poisson mixed model. The authors identified 1,031 DEGs using this combinatorial approach – DEGs requiring an FDR <0.01 in the cell-level and an FDR<0.05 in the patient level analysis. It is important to note that this cell-level differential expression approach, also known as pseudoreplication, over-estimates the confidence in DEGs due to the statistical dependence between cells from the same patient not being considered11,12,13,14. When we inspect all DEGs identified at an FDR of 0.05 from the authors’ cell-level analysis, this number increases to 14,274. Pseudobulk differential expression (DE) analysis has recently been proven to give optimal performance compared to both mixed models and pseudoreplication approaches11,12,15,16. It aggregates counts to individuals thus accounting for the dependence between an individual’s cells.

Here, to compare the effect of the different DE approaches in isolation, we apply a pseudobulk DE approach, sum aggregation and edgeR LRT17, to the authors original processed data. We found 26 unique DEGs when considering the six cell types used by the authors (Supplementary Table 2, Supplementary Fig. 3). This was 549 times fewer DEGs than that reported originally at an FDR of 0.05. When we compare these DEGs, we can see that the absolute log2 fold change (LFC) of our DEGs is 15 times larger than the authors’; median LFC of 2.34 and 0.16, despite the authors’ DEGs having an FDR score 8,000 times smaller; median FDR of 2.89x10-7 and 0.002 (Fig. 1a-b). Although we examined a high correlation in the genes’ fold change values across our pseudobulk analysis and the author’s pseudoreplication analysis (Pearson R of 0.87 for an adjusted p-value of 0.05, Supplementary Table 3), the p-values and resulting DEGs vary considerably. The correspondence in fold change values is expected given the approaches are applied to the same dataset whereas the probabilities, which pertain to the likelihood that a gene’s expressional changes is related to the case/control differences in AD, importantly do not align. We can show that this stark contrast is just an artefact of the authors taking cells as independent replicates and thus artificially inflating confidence by considering the Pearson correlation between the number of DEGs found and the cell counts (Fig. 1c-e). There is a near perfect, positive correlation between DEG and cell counts for the authors’ pseudoreplication analysis (Fig. 1c) and for the 1,031 genes from the authors’ combinatorial approach (Fig. 1d) which is not present in our pseudobulk re-analysis (Fig. 1e).

a, b highlight the log2 fold change and -log10 false discovery rate (FDR) of the differentially expressed genes from the author’s original work (Mathys et al.) and our reanalysis (Our analysis). In b, we have marked an FDR of 5x10-7, dashed grey line, to highlight how small the p-values from Mathys et al.’s analysis are. c, d, e, f, g show the Pearson correlation between the cell counts after QC and the number of DEGs identified. For f, g analysis, the samples have been randomly mixed between case and control patients. The different cell types are astrocytes (Astro), excitatory neurons (Exc), inhibitory neurons (Inh), microglia (Micro), oligodendrocytes (Oligo) and oligodendrocyte precursor cells (OPC).

A further point which questions the authors’ DE approach is that they identified the vast majority of DEGs in the more abundant, neuronal cell types1. However, an increase in the number of cells is not the same as an increase in sample size since these cells are not independent from one another - they come from the same sample. Therefore, an increase in the number of cells should not necessarily result in an increase in the number of DEGs whereas an increase in the number of samples would. This point is the major issue with pseudoreplication approaches which over-estimate confidence when performing differential expression due to the statistical dependence between cells from the same patient not being considered12,14. In our opinion, it makes more sense to identify the majority of large effect size DEGs in microglia which recent work has established is the primary cell type by which the genetic risk for Alzheimer’s disease acts18,19. This is what we found with our pseudobulk DE approach - 96% of all DEGs were in microglia (Supplementary Table 2) whereas only 3% of the authors’ DEGs were in microglia.

Although it has been proven that pseudoreplication approaches result in false positives by artificially inflating the confidence from non-independent samples, we wanted to investigate the effect of the approach on the authors’ dataset. We ran the same cell-level analysis approach – a Wilcoxon rank-sum test and FDR multiple-testing correction, 100 times whilst randomly permuting the patient identifiers (Fig. 1f). We would expect to find minimal DEGs with this approach given the random mixing of case and control patients. However, this pseudoreplication approach consistently found high numbers of DEGs and we observe the same correlation between the number of cells and number of DEGs as with the authors results. We did not observe the same pattern when running the same analysis with pseudobulk differential expression (Fig. 1g). As a result, we conclude that integrating this pseudoreplication approach with a mixed model like the authors proposed just artificially inflates the test confidence for a random sample of the genes resulting in more false discoveries in cell types with bigger counts.

Up to this point, to compare the effect of the DE approaches in isolation, we analysed the same processed data from the authors as opposed to our reprocessed data. We also performed pseudobulk DE on our reprocessed data and found 16 unique DEGs (Supplementary Table 4). It is worth noting that the fold change correlation between our two DE analyses (reprocessed data vs authors processed data) on the identified DEGs is only moderate (Pearson R of 0.57) and is lower than that of the correlation between pseudoreplication and pseudobulk on the same dataset (Supplementary Table 3). This highlights the effect that the low quality, high mitochondrial read cells have on DE analysis.

Conclusion

In conclusion, the authors’ analysis has been highly influential in the field with numerous studies undertaken based on their results, something we show has uncertain foundations. However, we would like to highlight that the use of pseudoreplication in neuroscience research is not isolated to the author’s work, others have used this approach20,21,22 and their results should be similarly scrutinised. Here, we provide our processed count matrix with metadata and also, the DEGs identified using an independently validated, differential expression approach so that other researchers can use this rich dataset free from spurious nuclei or DEGs. While the number of DEGs found here are significantly lower, much greater confidence can be had that these are AD relevant genes. The low number of DEGs found may also cause concern given the sample size and cost of collection and sequencing of such datasets. However, the increasing number of snRNA-Seq studies being conducted for AD, creates the opportunity to conduct differential meta-analyses to increase power. Further work is required in the field to develop methods to conduct such analysis, integrating studies and accounting for their the heterogeneity, similar to that which has been done for bulk RNA-Seq23. Some such approaches have already been made in COVID-19 research which could be leveraged for neurodegenerative disease24.

Supporting information